\chapter{Theoretical Fundamentals}
\label{c:fundamentals}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% ==========================================
% Section: Fundamentals of Quantum Information
% ==========================================
\section{Fundamentals of Quantum Information}
This section outlines the core principles of quantum information theory, establishing the essential theoretical vocabulary necessary for navigating the complexities of quantum computation.

We begin by defining the mathematical representation of quantum states. To describe composite systems consisting of multiple interacting components, we introduce the tensor product, which enables us also to represent the superpostions. Within this combined space, measure physical properties and defining interactions requires linear operators and observables. Finally, we detail the unitary operations and their underlying mathematical generators. Together, these foundational concepts provide the prerequisite knowledge required to understand the quantum formalisms discussed in subsequent chapters.

% ==========================================
% Subsection: Quantum States
% ==========================================
\subsection{Quantum States}
In an $n$-dimensional Hilbert space with a chosen orthonormal basis, a quantum state, denoted by the ket $\ket{\psi}$, can be represented as a $n$-dimensional complex column vector $\vec{\psi}$:
\begin{equation}
    \ket{\psi} \doteq \vec{\psi} =
    \begin{pmatrix}
        c_1 \\ c_2 \\ \vdots \\ c_n
    \end{pmatrix}.
\end{equation}

Its corresponding dual state, denoted by the bra $\bra{\psi}$, is represented by the conjugate transpose of this column vector, resulting in an $n$-dimensional row vector $\vec{\psi}^\dagger$:
\begin{equation}
    \langle\psi| \doteq \vec{\psi}^\dagger =
    \begin{pmatrix}
        c_1^* & c_2^* & \dots & c_n^*
    \end{pmatrix}.
\end{equation}


To ensure consistency with the probability interpretation, quantum states must be normalized to unity. Therefore, for a quantum state $\ket{\psi}$, the sum of the absolute squares of its probability amplitudes must equal to one:
\begin{equation}
    \abs{c_1}^2 +\abs{c_2}^2 + \dots + \abs{c_n}^2 = 1.
\end{equation}
\cite[chapter 1.1, 2.1]{Kottmann2025}
\cite[chapter 1.2.1]{sakurai2020modern}


% ==========================================
% Subsection: Tensor Product
% ==========================================
\subsection{Tensor Products}
% multi-particle system's are constructed in Hilbert space using tensor products (maybe)
In quantum mechanics, multi-particle systems are constructed within a composite Hilbert space using tensor products. The tensor product combines two or more distinct vector spaces into a single, larger space. Mathematically, if we consider a vector space of dimension $d_1$ and a second vector space of dimension $d_2$, the resulting composite space has a dimension of $d_1 \cdot d_2$ (i.e., $\mathbb{C}^{d_1} \otimes \mathbb{C}^{d_2} \rightarrow \mathbb{C}^{d_1 \cdot d_2}$).

Constructing a multi-qubit basis state is achieved by taking the tensor product of the individual single-qubit states. For instance, if one qubit is in the state $\ket{0}$ and a second is in the state $\ket{1}$, the joint state of the two-qubit system is $\ket{0} \otimes \ket{1}$, which is conventionally abbreviated as $\ket{01}$. Because quantum systems can exist in superposition, a general composite state is expressed as a linear combination of all possible tensor product basis states. For a two-qubit system, this yields a four-dimensional complex vector:
\begin{equation}
    \ket{\psi} = c_{00}\ket{00} + c_{01}\ket{01} + c_{10}\ket{10} + c_{11}\ket{11}.
\end{equation}

The tensor product formalism also provides a definition of entanglement. A multi-particle state is defined as entangled if it cannot be factored into a simple tensor product of individual, single-particle states (i.e., $\ket{\psi} \neq \ket{\phi} \otimes \ket{\varphi}$).


When quantum operators act on these composite Hilbert spaces, their tensor products obey several fundamental algebraic properties:
\begin{itemize}
    \item \textbf{Non-commutativity:} The order of the tensor product strictly matters; in general, $X \otimes Y \neq Y \otimes X$.
    \item \textbf{Bilinearity:} The operation distributes over addition, such that $X \otimes (Y + Z) = X \otimes Y + X \otimes Z$.
    \item \textbf{Mixed-Product Property:} Operators act independently on their respective subsystems, following the rule $(A \otimes B)(C \otimes D) = AC \otimes BD$.
    \item \textbf{Eigensystems:} If an operator is formed by the tensor product of two matrices, its eigenvectors are the tensor products of the individual eigenvectors ($\ket{a_k} \otimes \ket{b_l}$), and the corresponding eigenvalues are the products of the individual eigenvalues ($a_k \cdot b_l$).
\end{itemize}

\cite[chapter 2.3, A.8]{Kottmann2025}



% ==========================================
% Subsection: Operators and Observables
% ==========================================
\subsection{Operators and Observables}
A linear operator acts on a ket from the left to produce another ket vector, or on a bra from the right to produce another bra vector:
\begin{align}
    X\ket{\alpha} &= \ket{\beta}, \\
    \bra{\alpha}X &= \bra{\beta}.
\end{align}

An operator $X$ is defined as the \textbf{null operator} if, for any arbitrary ket $\ket{\alpha}$, it satisfies:
\begin{equation}
    X \ket{\alpha} = 0.
\end{equation}

Operators can be added ($X + Y = Y + X$) and multiplied. While operator multiplication is generally noncommutative ($XY \neq YX$), it is strictly associative:
\begin{equation}
    X(YZ) = (XY)Z = XYZ.
\end{equation}


Because quantum states are represented as vectors in $\mathbb{C}^n$, a linear operator can be represented by a complex $n \times n$ matrix $U$. Given an input ket $\ket{\psi}$, the resulting state $\ket{\phi}$ is calculated via matrix-vector multiplication:
\begin{equation}
    \ket{\phi} = U\ket{\psi}.
\end{equation}

Crucially, physical operations that evolve a quantum state must preserve its probability normalization. Consequently, physical evolution operators are restricted to unitary matrices, ensuring that the inner product—and thus the normalization—remains invariant

An \textbf{observable} represents a physical quantity of a system that can be experimentally measured. Classical examples include position, momentum, angular momentum, and energy. In an $N$-qubit system, an observable is defined mathematically as a $2^N \times 2^N$ Hermitian matrix $H$. Because the operator is Hermitian, it guarantees that all possible measurement outcomes are strictly real numbers and correspond directly to the eigenvalues of $H$. Crucially, the act of measurement fundamentally alters the system: upon obtaining a specific eigenvalue, the quantum state instantaneously collapses into the corresponding eigenstate of that observable

Due to the inherent probabilistic nature of quantum mechanics, we generally cannot predict the outcome of a single measurement with absolute certainty. Instead, we evaluate the \textbf{expectation value}, which represents the statistical average of measured values over a large ensemble of identically prepared systems

In the context of this thesis, the primary observables of interest are Hamiltonians, which represent the total energy of the physical system. Let $H$ be the Hamiltonian. We conventionally assume that its resulting energy eigenvalues are ordered from the ground state upwards ($E_0 \le E_1 \le E_2 \le \dots$), satisfying the eigenvalue equation:
\begin{equation}
    H \ket{E_i} = E_i \ket{E_i}.
\end{equation}

Given a hamiltonian Operator $H$ and a quantum state $\ket{\psi}$ the expectation value can be calculated as:
\begin{equation}
    \langle H\rangle _{\psi} \coloneqq \bra{\psi} H \ket    {\psi}.
\end{equation}
\cite[chapter 2.2, 5.3]{Kottmann2025}
\cite[chapter 1.2, 1.4.1]{sakurai2020modern}
\cite[chapter 1.1.1]{schleich2025quantum}

% ==========================================
% Subsection: Unitary Operations and Generators
% ==========================================
\subsection{Unitary Operations and Generators}
The physical transformations applied to quantum states are represented by unitary operators. To construct parameterized quantum gates, these unitary operations are typically expressed via the exponential map:
\begin{equation}
    \label{unitary_operation}
    U(\theta) = e^{-i\frac{\theta}{2}G},
\end{equation}
where $\theta$ is a real-valued scalar parameter (such as a time or rotation angle), and $G$ is a Hermitian matrix ($G = G^\dagger$) known as the generator of the operation.
\cite[chapter 1.2.12]{benito_kottmann_qc_2025}
\cite[chapter II.]{kottmann2020feasible}



% ==========================================
% Section: Fermionic Many-Body Systems
% ==========================================
\section{Fermionic Many-Body Systems}
Fermions are fundamental particles such as electrons, protons, and neutrons, constitute the primary building blocks of matter. In the context of computational many-body physics, the challenge of simulating these systems lies not in their isolated particle properties, but in the collective quantum mechanical rules governing their many-body dynamics. Specifically, a simulation must describe the rules of fermions and their interactions accurately.
\cite[chapter 6.1]{benito_kottmann_qc_2025}

% ==========================================
% Subsection: Indistinguishability
% ==========================================
\subsection{Indistinguishability}
In classical physics, identical objects can be tracked along distinct, observable trajectories. In quantum mechanics, however, identical particles are fundamentally indistinguishable; it is physically impossible to track their individual paths or assign them persistent labels without collapsing or disturbing the system's state. Because measurable physical quantities are calculated as expectation values, which rely on the product of the wavefunction and its complex conjugate ($\bra{\psi} H \ket{\psi} = \sum_{i}{\abs{c_i}^2} E_i$), the indistinguishability principle dictates that these observables must remain strictly invariant under the permutation (exchange) of any two identical particles.
\cite[chapter 7.1, 7.2]{sakurai2020modern}
\cite[chapter 6.1]{benito_kottmann_qc_2025}


% ==========================================
% Subsection: Wavefunction Symmetry and Antisymmetry
% ==========================================
\subsection{Wavefunction Symmetry and Antisymmetry}
In many-body quantum mechanics, systems of identical particles are tightly constrained by symmetry. The total wavefunction of such a system must be either entirely symmetric or entirely antisymmetric under the exchange of any two identical particles.

\textbf{Antisymmetric Wavefunctions:} Fermions are strictly governed by totally antisymmetric wavefunctions. Consequently, exchanging the spatial and spin coordinates of any two identical fermions introduces a global phase factor of $-1$ to the wavefunction. The Pauli Exclusion Principle is a direct mathematical corollary of this antisymmetry; if two identical fermions were to occupy the identical quantum state, their exchange would require the wavefunction to equal its own negative, implying the wavefunction must identically vanish.

\textbf{Symmetric Wavefunctions:} Bosons, conversely, are described by totally symmetric wavefunctions. Exchanging any two identical bosons yields a phase factor of $+1$, leaving the wavefunction fundamentally unaltered. Free from the antisymmetric cancellation constraints that restrict fermions, identical bosons can condense into the same quantum state.

\cite[chapter 7.1, 7.2]{sakurai2020modern}
\cite[chapter 6.1]{benito_kottmann_qc_2025}

% ==========================================
% Subsection: Pauli Exclusion Principle
% ==========================================
\subsection{Pauli Exclusion Principle}
The Pauli Exclusion Principle, formulated by Wolfgang Pauli in 1925, is a direct physical consequence of the indistinguishability applied specifically to fermions. It establishes that no two identical fermions can simultaneously occupy the exact same quantum state within a given system. Originally conceptualized to explain atomic electron configurations, where no two electrons can share the same four quantum numbers, this principle is a defining, non-negotiable characteristic of fermionic behavior. This exclusion must be preserved. Any application of creation ($a_i^\dagger$) or annihilation ($a_i$) operators must naturally prevent multiple occupancies. \cite[chapter 7.1, 7.2]{sakurai2020modern}

% ==========================================
% Section: The Second Quantization Framework
% ==========================================
\section{The Second Quantization Framework}
In the first quantization formalism, describing a many-body fermionic system requires the explicit construction of fully antisymmetric wavefunctions. As the number of particles increases, tracking the coordinates of individual indistinguishable particles becomes hard to deal. The second quantization framework fundamentally resolves this bottleneck. Rather than defining explicit wavefunctions for individual particles, it shifts the perspective to the occupation of predefined single-particle states within a Fock space. Particle indistinguishability and wavefunction antisymmetry are not discarded, they are rather embedded directly into the anticommutation algebra of the fermionic creation and annihilation operators. This provides a scalable, operator-driven foundation.

\cite[chapter 2.1]{schleich2025quantum}
\cite[chapter 6.5]{benito_kottmann_qc_2025}
\cite[chapter 7.7.1]{sakurai2020modern}

% ==========================================
% Subsection: Fock Space
% ==========================================
\subsection{Fock Space}
To mathematically formalize the second quantization framework, we require a state space capable of accommodating a variable number of particles. This is achieved by constructing the Fock space, $\mathcal{F}$, which is defined as the direct sum of $N$-particle Hilbert spaces. This structure allows us to express states without being restricted to a fixed particle number.

Because this space utilizes the second quantization formalism, it does not track the explicit spatial coordinates of each particle. Instead, states in the Fock space are expressed using the occupation number representation:
\begin{equation}
    \ket{\psi} = \ket{n_1, n_2, \dots, n_i, \dots}.
\end{equation}
In this notation, each index $i$ corresponds to a specific single-particle state (or spin-orbital), and $n_i$ denotes its respective occupation number. Note that due to the Pauli exclusion principle, the occupation number for fermions is strictly restricted to $n_i \in \{0, 1\}$.

To illustrate this representation, we can define several fundamental states. The vacuum state, representing a system empty system without any particles, is denoted as:
\begin{equation}
    \ket{0, 0, \dots, 0, \dots} = \ket{\boldsymbol{0}}.
\end{equation}

A single-particle state, where only the $i$-th spin-orbital is occupied, is written by setting $n_i = 1$:
\begin{equation}
    \ket{0, \dots, n_i=1, \dots} = \ket{k_i}.
\end{equation}

More complex, multi-particle states are systematically generated by applying creation and annihilation operators to these basis states. For example in a system with 4 spin-orbitals, the occupation vector $\ket{0110}$  indicates that the second and third spin-orbitals are occupied by particles, while the first and fourth remain empty.

% ==========================================
% Subsection: Fermionic Operators
% ==========================================
\subsection{Fermionc Operators}
In quantum mechanics, a many-body system of identical particles is described using the formalism of second quantization. Within this framework, the physical state of a system is manipulated via fundamental mathematical tools known as creation ($\hat{a}_k^\dagger$) and annihilation ($\hat{a}_k$) operators. These operators act directly upon a quantum state: $\hat{a}_k^\dagger$ creates a fermion in a specific spin-orbital $k$, while $\hat{a}_k$ destroys a fermion in that same orbital. Because fermions inherently obey the Pauli Exclusion Principle, these operators naturally enforce strict physical boundaries; applying a creation operator to an already occupied orbital, or an annihilation operator to an empty one, identically annihilates the wavefunction (yielding zero). These characteristics can be noted down as:
\begin{align}
    \hat{a} |0\rangle &= 0, \\
    \hat{a}^\dagger |0\rangle &= |1\rangle, \\
    \hat{a} |1\rangle &= |0\rangle, \\
    \hat{a}^\dagger |1\rangle &= 0.
\end{align}

\cite[chapter 2.1]{schleich2025quantum}


% ==========================================
% Section: Mapping Fermions to Quantum Hardware
% ==========================================
\section{Mapping Fermions to Quantum Hardware}
The objective of a fermionic simulator is to evaluate the dynamics of many-body fermionic systems using qubit-based architectures. However, quantum hardware exclusively executes operations on distinguishable qubits. Consequently, a mathematically encoding scheme is required to translate fermionic behavior into a native qubit representation. This section first examines the fundamental algebraic mismatch between fermions and qubits, and subsequently introduces the specific mapping mechanisms to bridge this hardware gap.

% ==========================================
% Subsection: The Hardware Constraint: Fermions vs. Qubits
% ==========================================
\subsection{The Hardware Constraint: Fermions vs. Qubits}

While fermions are physical particles governed by indistinguishability, qubits are idealized, fully distinguishable two-state systems that serve as the fundamental units of quantum computation. In a quantum processor or qubit-based simulator, individual qubits are independently addressable. This distinguishability yields a crucial algebraic divergence: operators acting on disparate qubits naturally commute:
\begin{equation}
    [A, B] = AB - BA = 0 \quad \Rightarrow \quad AB = BA
\end{equation}
In contrast, fermionic creation ($\hat{a}_i^\dagger$) and annihilation ($\hat{a}_i$) operators must anticommute:
\begin{equation}
    \{\hat{a}_i^\dagger, \hat{a}_j^\dagger\} = \hat{a}_i^\dagger \hat{a}_j^\dagger + \hat{a}_j^\dagger \hat{a}_i^\dagger = 0 \quad \Rightarrow \quad \hat{a}_i^\dagger \hat{a}_j^\dagger = -\hat{a}_j^\dagger \hat{a}_i^\dagger
\end{equation}
This chrectersitc is needed to preserve wavefunction antisymmetry and enforce the Pauli Exclusion Principle.

\cite[chapter 3.11.1, 7.1, 7.7]{sakurai2020modern}
\cite[chapter 1.]{Kottmann2025}


% ==========================================
% Subsection: The Jordan-Wigner Transformation
% ==========================================
\subsection{The Jordan-Wigner Transformation}
To simulate fermionic systems on quantum hardware, an encoding scheme is required to map the physical particles to the computational framework. The Jordan-Wigner (JW) transformation serves as the foundational mechanism for this mapping, establishing a direct bijection between $N$ fermionic spin-orbitals and $N$ individual qubits. Within this framework, the computational basis states directly represent the Fock space occupation number vectors: a qubit in state $\ket{1}$ denotes an occupied spin-orbital, while $\ket{0}$ denotes an empty one.

However, preserving the indistinguishability of fermions presents a significant hardware constraint. Fermionic creation and annihilation operators are mathematically constrained to strictly anticommute to guarantee that the many-particle wavefunction remains totally antisymmetric under particle exchange. Because quantum hardware utilizes distinguishable qubits whose operators naturally commute, fermionic operators cannot be mapped to single qubits in a simple one-to-one fashion. Instead, they must be explicitly constructed from composite qubit operators.

Under the standard Jordan-Wigner orbital-to-qubit mapping, the fermionic operators are defined as:
\begin{align}
    \label{eq:creation_op_jd_transformation}
    \hat{a}_k^\dagger &= \mathbb{I}^{\otimes (k-1)} \otimes \hat{\sigma}_k^+ \otimes \hat{\sigma}_z^{\otimes (N-k)}, \\
    \label{eq:annhilation_op_jd_transformation}
    \hat{a}_k &= \mathbb{I}^{\otimes (k-1)} \otimes \hat{\sigma}_k^- \otimes \hat{\sigma}_z^{\otimes (N-k)}.
\end{align}
This composite definition relies on two functional components to artificially replicate fermionic behavior within an inherently commuting qubit environment:

\textbf{Ladder Operators ($\hat{\sigma}_k^\pm$):} Defined using the standard Pauli matrices as:
\begin{equation}
    \hat{\sigma}_k^\pm = \frac{1}{2}(\hat{\sigma}_x \pm i\hat{\sigma}_y).\label{eq:ladder_operators}
\end{equation}
These operators execute the local ``creation'' or ``annihilation'' at the specific target qubit $k$ by raising or lowering the computational basis state between $\ket{0}$ (empty) and $\ket{1}$ (occupied).

\textbf{Parity Strings ($\hat{\sigma}_z^{\otimes (N-k)}$):} A non-local tensor product of Pauli-$Z$ operators acting on the adjacent qubits. This string encodes the necessary parity information into the phase of the wavefunction.

When evaluated against a quantum state, these non-local parity strings ($\hat{\sigma}_z$) systematically scan all qubits succeeding (or preceding, depending on the chosen convention) the target site $k$. Because the Pauli-$Z$ operator acts as the identity on $\ket{0}$ and applies a phase of $-1$ to $\ket{1}$, this string iteratively multiplies a global phase factor of $-1$ for every occupied state it encounters. This mechanism systematically and globally replicates the exact antisymmetric phase inversions demanded by the Pauli Exclusion Principle.

\cite[chapter 2.1]{schleich2025quantum}
\cite[chapter III.B]{anand2022quantum}
\cite[chapter II.E]{kottmann2020feasible}
\cite[chapter 1.1, 5.3]{Kottmann2025}


% ==========================================
% Section: Simulating Excitations
% ==========================================
\section{Simulating Excitations}
An excitation in the quantum mechanic describes the transition of a particle such as an electron from a lower energy orbital to a higher energy orbital. We can use this phänomena to move particles between the orbitals and therefore manipulate our current state of the system. This is gonna be our goal to achieve in our simulator.

Because this principle works different depending on if the system we work with is a qubit or a fermionic one, we gonna discuss the differences in this section. In general we use for both excitations a rotation operator and change the Generator of each one.

% ==========================================
% ==========================================
\section{Simulating Excitations}
In quantum mechanics, an excitation describes the transition of a particle, such as an electron, from a lower-energy orbital to a higher-energy state. Within the context of computational many-body physics, these phenomena can manipulate the quantum state of the system. Implementing and evaluating these operations efficiently is the primary objective of the simulator.

We saw before that fermions differ from qubits, also in the defintion of the excitation of the both particle groups differences can be seen. In both cases, excitations are enacted via unitary evolution operators; however, the explicit construction of the underlying mathematical generators must be changed to obey the respective commutation or anticommutation rules of the system. This is detailed in this section.


% ==========================================
% Subsection: Qubit Excitations
% ==========================================
\subsection{Qubit Excitations}
To dynamically evolve a quantum state on hardware, the most direct approach is to apply excitations directly within the computational qubit space. These operations, commonly referred to as qubit excitations or orbital rotations, move particle occupations between sites. Instead of utilizing true fermionic creation ($\hat{a}^\dagger$) and annihilation ($\hat{a}$) operators, qubit excitations are constructed natively from the Ladder Operators.

Mathematically, a single-qubit excitation describing a transition between an occupied qubit $i$ and a target qubit $b$ is defined by the unitary operator:
\begin{equation}
    U_{i}^{b}(\theta) = e^{\theta \left(\hat{\sigma}^+_{b}\hat{\sigma}^-_{i} - \hat{\sigma}^+_{i}\hat{\sigma}^-_{b}\right)}.\label{eq:qubit_exc_unitary_def}
\end{equation}
This can be equivalently expressed in terms of standard Pauli matrices (absorbing the scalar factors into the rotation parameter $\theta$) as (see appendix for algebraically derivation):
\begin{equation}
    U_{i}^{b}(\theta) = e^{-i\theta \left(\hat{\sigma}^b_{x}\hat{\sigma}^i_{y} - \hat{\sigma}^b_{y}\hat{\sigma}^i_{x}\right)}.\label{eq:quibt_exc_unitary_derivation}
\end{equation}
Note that $\hat{\sigma}_{x}$ and $\hat{\sigma}_{y}$ refers to the Pauli-X and Pauli-Y operators. This unitary operator acts locally on the target qubits, functioning as a rotation matrix exclusively within the $\{\ket{01}, \ket{10}\}$ subspace to swap the occupation states.

\cite[Chapter IV.A.1.b.]{anand2022quantum}

This concept naturally extends to higher-order excitations. Given a set of target orbitals denoted by the vector $\boldsymbol{k}$, and a set of occupied orbitals denoted by $\boldsymbol{l}$, the multi-qubit excitation generator is constructed directly from products of the corresponding raising and lowering operators:
\begin{equation}
    \label{eq:mult_qu_exc_generator}
    G^{\boldsymbol{k}}_{\boldsymbol{l}} = i\left(\prod_n \hat{\sigma}^-_{k_n} \prod_m \hat{\sigma}^+_{l_m} - \text{h.c.} \right).
\end{equation}
% Todo reference

% ==========================================
% Subsection: Fermionic Excitations
% ==========================================
\subsection{Fermionic Excitations}
While qubit excitations are highly hardware-efficient and successfully redistribute particle occupations, they lack a crucial physical property: they do not natively track the non-local antisymmetric phase shifts required when identical fermions exchange positions. To fully capture the physical reality of a many-body system and enforce the Pauli Exclusion Principle globally, we must utilize true fermionic excitations.

Fermionic excitations are represented by a generator $G$, constructed strictly from the anticommuting creation ($\hat{a}^\dagger$) and annihilation ($\hat{a}$) operators introduced during second quantization. For instance, the generator $G_{pq}$ describes a single fermionic excitation, denoting the transition of an electron from the $p$-th spin-orbital to the $q$-th spin-orbital:
\begin{equation}
    G_{pq} = i(\hat{a}^\dagger_p \hat{a}_q - \hat{a}^\dagger_q \hat{a}_p).
\end{equation}

Higher-order excitations are constructed in a similar mathematical manner. The generator $G_{pqrs}$ models a double excitation, accounting for the simultaneous transition of one electron between the $p$-th and $q$-th orbitals, and a second electron between the $r$-th and $s$-th orbitals:
\begin{equation}
    G_{pqrs} = i(\hat{a}^\dagger_p \hat{a}_q \hat{a}^\dagger_r \hat{a}_s - \text{h.c.}).
\end{equation}
Here, $\text{h.c.}$ denotes the Hermitian conjugate of the preceding term. This standard notation simplifies the equation while ensuring the operator correctly captures the reverse electron exchange process to maintain Hermiticity, analogous to the subtracted term in the single excitation.

Generalizing this concept, an $n$-fold fermionic excitation can be compactly represented by a generator $G_{\mathbf{pq}}$, which encapsulates all possible simultaneous electron exchanges between a set of occupied orbitals $\{p_k\}$ and a set of target orbitals $\{q_k\}$. Because this represents $n$ simultaneous events, it is defined as a product of $n$ operator pairs:
\begin{equation}
    \label{eq:mult_fe_exc_generator}
    G_{\mathbf{pq}} = i \left( \prod_{k=1}^{n} \hat{a}^\dagger_{p_k} \hat{a}_{q_k} - \text{h.c.} \right).
\end{equation}
To apply these generators within a quantum simulator or on physical quantum hardware, they are mapped via the Jordan-Wigner transformation and exponentiated into a unitary operation (e.g., $U = \exp(-i\theta G_{\mathbf{pq}})$).

\cite[chapter II.]{kottmann2020feasible}\cite[chapter III.A]{anand2022quantum}




















