\chapter{Implementation in C++}
\label{c:implementation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ==========================================
% Section: Core Data Structures
% ==========================================
\section{Core Data Structures}
To construct the foundation of the simulator, we must first define the core data structures that map the mathematical framework into efficient computational memory.

% ==========================================
% Section: Defining the state
% ==========================================
\subsection{Defining the state}
As established in Sections~\ref{subsec:quantum-states} and~\ref{subsec:tensor-products}, a quantum state is represented as a linear combination of orthogonal basis states $\ket{i}$ and their corresponding complex probability amplitudes $c_i$. Computationally, simulating these many-body states requires a data structure optimized for rapid access and modification. This is particularly crucial because, during generator-based evolution, the algorithm must dynamically retrieve and manipulate specific pairs of interacting basis states.

To represent the fermionic Fock states, we use the occupation number representation defined in Equation \eqref{eq:occupation_number_representation}. Because fermions obey the Pauli exclusion principle, a given orbital is either strictly empty or occupied. This binary nature maps perfectly to a bitstring. In C++, this is naturally and efficiently encoded as an unsigned 64-bit integer (\texttt{uint64\_t}). Utilizing native integer types allows the simulator to execute optimized, hardware-level bitwise operations when applying creation and annihilation operators. For the associated probability amplitudes, we utilize the standard library's double-precision complex type, \texttt{std::complex<double>}, ensuring robust numerical stability during phase accumulations.

To bind the basis states to their amplitudes, we require a structure that supports dynamic sizing and fast lookups. Rather than relying on dense arrays—which would scale exponentially and store unnecessary zero-amplitude states—we implement a sparse representation using a hash map. Specifically, we utilize the \texttt{ankerl::unordered\_dense::map}. Unlike the standard \texttt{std::unordered\_map}, this optimized library provides a cache-friendly, contiguous memory layout.

Combining these architectural choices, the complete quantum state is elegantly defined via a single type alias:

\begin{minted}{cpp}
    using State = ankerl::unordered_dense::map<uint64_t, std::complex<double>>;
\end{minted}

% ==========================================
% Section: Excitations
% ==========================================
\section{Excitations}
Building upon the core state representation, we now address the application of excitation operators. We first examine the simplified qubit excitation baseline before introducing the necessary fermionic modifications.

% ==========================================
% Subsection: Qubit Baseline
% ==========================================
\subsection{Qubit Baseline}
As discussed in Section~\ref{subsec:qubit-baseline}, implementing the qubit excitation requires several key parameters: the initial quantum state, the orbital sets $\boldsymbol{k}$ and $\boldsymbol{l}$ (which define the electron transitions), and the rotation angle $\theta$.

Computationally, the orbital sets are passed as constant references to standard integer vectors (\texttt{std::vector<int>}) to avoid unnecessary memory allocations, while the angle $\theta$ is provided as a double-precision float. This results in the following function signature:

\begin{minted}{cpp}
State apply_qubit_excitation(const State& state, const std::vector<int>& k, const std::vector<int>& l, const double theta)
\end{minted}

The core algorithm must iterate over all non-zero basis states and apply the unitary transformation defined in Equation \eqref{eq:qubit-excitation-unitary-transformation-configuration-differences}:

\begin{equation} \tag{\ref{eq:qubit-excitation-unitary-transformation-configuration-differences}}
    \qubitExcitationUnitaryTransformationConfigurationDifferences
\end{equation}

To iterate efficiently over the sparse state map, we utilize C++17 structured bindings alongside the \texttt{auto} keyword. This modern language feature allows direct, reference-based extraction of the basis state and its corresponding amplitude without invoking costly data copying:

\begin{minted}{cpp}
    for (const auto& [basis_state, coeff] : state)
\end{minted}

Before performing any computationally expensive floating-point arithmetic or creating new state entries, the simulator evaluates the nullspace projector $P_0$. A basis state belongs to the active space (avoiding the nullspace) if and only if all origin orbitals $\boldsymbol{k}$ are uniformly occupied while all target orbitals $\boldsymbol{l}$ are uniformly unoccupied, or vice versa.

Because the basis state is elegantly encoded as a \texttt{uint64\_t}, we can evaluate this physical condition using raw hardware-level bitwise operations. Bit shifting (\texttt{>>}) and masking (\texttt{\& 1ULL}) execute in a single CPU cycle. By looping over the orbital indices, the algorithm can aggressively prune nullspace states with virtually zero computational overhead:

\begin{minted}{cpp}
    bool k_val = (basis_state >> k[0]) & 1ULL;
    bool skip = false;
    for (size_t i = 0; i < k.size(); i++) {
        if (((basis_state >> k[i]) & 1ULL) != k_val) skip = true;
        if (((basis_state >> l[i]) & 1ULL) == k_val) skip = true;
    }
\end{minted}

We now know that the basis state is in the active space so we can perform the calculations of the excitation. We again already discussed in section~\ref{subsec:qubit-baseline} how the phases needed to get updated. Which we can easily apply in C++. First the phase of the basis state gets updated as in equation \eqref{eq:qubit-excitation-update-basis-state}:
\begin{equation} \tag{\eqref{eq:qubit-excitation-update-basis-state}}
    c_{\boldsymbol{i}} \leftarrow c_{\boldsymbol{i}} \cos\left(\frac{\theta}{2}\right)
\end{equation}

